Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO: a Bayesian study

### 2016-05-04 Paper/paper.tex
notes tex
making changes from comments

\author{Prayush Kumar}\CITA\email{prkumar@cita.utoronto.ca}
\author{Michael P\"urrer}\AEI
\author{Harald P. Pfeiffer}\CITA\AEI

@@

\begin{abstract}
Very recently, the Advanced LIGO (aLIGO) observatories announced the first
terrestrial detection of gravitational waves. This heralds an era of 
observational gravitational wave (GW) astronomy. Neutron star - black hole
(NSBH) binaries are one of the primary sources targeted for observation
with the aLIGO detectors.
% 
GWs from these sources will carry signatures of the tidal distortion of
the NS by the BH during the inspiral, and of its tidal disruption (if that
happens) at merger. In the NS is disrupted, the merger and ringdown signal
is drastically suppressed, which distinguishes it from a BHBH merger.
% 
In this paper, we present the first Bayesian study of the measurability of
the tidal deformability of neutron stars, allowing the companion hole to take 
on a variety of masses and spins.
% 
First, we find that if BHBH templates are used to estimate source 
parameters for an NSBH signal, a significant bias is introduced in the 
inference if the signal-to-noise ratio (SNR) is greater than $\rho\simeq30$.
% 
Second, we find that if we are to get a loud signal with $\rho\geq 23$, 
we can put a factor of two bound on the NS tidal deformability parameter
$\lambdans\propto (R/M)^{5}$.
% 
Finally, we study the improvement in our measurement of $\lambdans$ from 
multiple observations. Using an approximately astrophysical population, 
we find that (i) the median measured $\lambdans$ catches up to within
$10\%$ of the true value in approximately $20$ observations; (ii) the $90\%$
confidence interval for $\lambdans$ shrinks to $\pm 50\%$ around the true 
value with the same number of observations.
% 
We find these results encouraging, and recommend that an effort to 
measure $\lambdans$ be taken within the LIGO-Virgo Collaboration once
we begin to detect NSBH coalescences.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{s1:introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The Advanced LIGO (aLIGO) observatories began their first observing run ``O1''
mid-September 2015, operating at a factor of $3-4$ higher gravitational-wave
(GW) strain sensitivity than their first-generation 
counterparts~\cite{Shoemaker2009}.
% 
During O1, the first terrestrial detection of gravitational waves was made, as
was announced earlier in 2016~\cite{LIGOVirgo2016a}. Emitted more than a
billion year ago, these waves herald the era of observational gravitational
wave (GW) astronomy as they pass through Earth.
% 
Over the $3$-months of O1, aLIGO had access to more binary mergers than 
initial-LIGO over its entire coincident operation period. Both of aLIGO 
instruments will be further upgraded in phases to reach their design 
sensitivity by $2018-19$, at which point we may expect to observe $\sim 70$ 
binary mergers a year~\cite{Abadie:2010cf}.

Coalescing binaries of compact objects, i.e. black holes (BH) and/or neutron 
stars (NS), remain the most interestings targets for GW based observation with
aLIGO. NSBH binaries form an important sub-class of compact binaries, and we
expect to observe $\mathcal{O}(10)$ of these every year with design 
sensitivity aLIGO~\cite{Abadie:2010cf}.
NSBH mergers are of interest for various reasons. For instance, they have 
been associated with (as the progenitors of) short Gamma-ray Bursts
(GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Under favorable conditions, close to merger the neutron star
can get disrupted by the tidal field of its companion hole. Once disrupted,
the neutron star material may form an accretion disk around the black hole. 
Most of the disk matter spirals in within a few milliseconds, with a small 
fraction becoming unbounded and getting ejected, with a possibled simultaneous
burst of gamma-ray emission)~\cite{Foucart2015,Lovelace:2013vma,Deaton2013,
Foucart2012,Shibata:2005mz,Paschalidis2014}. A coincident detection of GWs
from an NSBH system with a GRB would therefore provide us insights into the
central engine powering GRBs. 
Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron stars. This is what we will focus on in this paper.
The gravitational-waves emitted by NSBH binaries carry subtle hints of the 
matter structure of the neutron star. During early inspiral, the (weak) tidal
field of the companion hole deforms the star, thereby exciting its resonant
oscillation modes. The induced quadrupolar moment of the star depends on the
tidal field $\mathcal{E}_{ij}$ due to its companion through the relation
$Q_{ij} = -\lambda\mathcal{E}_{ij}$, where $\lambda$ is the EoS dependent
tidal deformability parameter related to the neutron star's dimensionless
Love number $k_2$ and radius $R_\mathrm{NS}$ as
$\lambda:=\frac{2}{3G}k_2 R_\mathrm{NS}^5$. This drains energy from the orbit
resulting in an increase of the inspiral rate. We know the leading and next-to
-leading order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN order. Close to
merger, as the NS gets disrupted and forms long tails that feed into the 
companion black hole, the quadrupole moment of the system falls monotonically.
As a result, the emitted GWs are subdued significantly post-disruption
(as compared to what they would be for a BBH),
and the quasi-normal modes of the resulting black hole are not 
excited~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,Foucart2015,
Pannarale:2015jia}.
The difficulty in discerning NS matter effects in an NSBH coalescence from its
GW signal lies in distinguishing it reliably from a BBH signal, since the 
difference between the two is either subtle (during inspiral) or short-lived
(disruption and merger). The earlier in inspiral the NS disrupts, the more
orbits there are over which its effect is visible. The mechanism of disruption
is governed by the nature of the BH, i.e. its mass and spin. More massive BHs 
(with $m_\mathrm{BH}\gtrsim 7\, m_\mathrm{NS}$), as well as BHs with
(large) retrograde spins $(\chi_\mathrm{BH}\in[-1. 0])$, tend to swallow the 
NS whole due to a rapid merger (i.e. without it disrupting before the
inner-most stable circular orbit (ISCO) is reached). The quasi-normal
oscillation modes of the final black hole's horizon are also 
excited~\cite{Foucart:2013psa}. Therefore, the only available information 
comes from the inspiral rate modification. On the other hand, less massive
black holes spinning in the prograde sense (with mass-ratio 
$q\in[2, 5]$, and $\chi_\mathrm{BH}\geq 0$) can cause their companion star to 
disrupt much before merger, leaving a strong imprint on the emitted GW 
signal~\cite{Shibata:2007zm,2010PhRvD..81f4026F,Lovelace:2013vma,
Kawaguchi:2015}.

@@

We first probe the effect of not including tidal effects in parameter 
estimation templates on the recovery of non-tidal binary parameters, such as 
component masses and spins. This is the case of current and planned aLIGO
efforts.
To do so, we use the LEA+ model to generate a set of realistic signals, and
use that as data from which we estimate binary mass and spin parameters with
a Bayesian Monte-Carlo algorithm, using non-tidal (BBH) waveforms as filters.
% 
As in rest of the paper, we restrict here our parameter space to span 
mass-ratios $q:=\mbh/\mns\in[2,5]$, black hole spin $\chibh\in[-0.5, +0.75]$, 
and dimensionless neutron star tidal deformability 
$\lambdans:= G\left(\frac{c^2}{G \mns}\right)^5\lambda \in[500, 2000]$.
The mass of NS for signals (and not templates) is fixed to $1.35M_\odot$ 
throughout~\cite{stellarcollapsemass}, and is taken to be 
non-spinning~\cite{Miller:2014aaa}.
Each signal is rescaled to a range of SNR values $\rho\in[10, 70]$.
% 
We find that non-inclusion of tidal effects leads to systematic biases in
the measurement of non-tidal parameters. Consider first the binary chirp 
mass $\mchirp:=M\eta^{3/5}$ where $M$ is the binary total mass, and 
$\eta:=\mbh\mns/M^2$ is the dimensionless mass-ratio. We find that for
realistic (moderate) SNRs, i.e. $\rho\lesssim 30$, not including tidal
effects introduces a barely measurable bias that remains below $50\%$
of the statistical uncertainty of the $\mchirp$ measurement itself.
For higher signal strengths, we find that the underlying statistical
uncertainty decreases enough for the systematic bias to become comparable
to it for $\rho=50-70$ and higher. This is expected because chirp mass is fairly
accurately determined by the inspiral, where tidal effects are weak.
Next, we consider the mass-ratio $\eta$. We find that only if the NS is quite
deformable (with $\lambdans=2000$), {\it and} has a companion BH whose mass
is $\mbh\lesssim 4.5M_\odot$ (i.e. in the astrophysical ``mass-gap''~\cite{
Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa}),
will the systematic bias in $\eta$ measurement become comparable to its
statistical uncertainties at SNRs $\rho\leq 30$. For SNRs $\rho\gtrsim30$,
however, not including tidal effects is very likely to {\it significantly} bias
the mass-ratio estimate for NSBH binaries. Finally, we consider the spin on the
black hole $\chibh$. We find that $\chibh$ measurements are affected in a similar
way to $\eta$. By not including tidal effects, we compromise our measurement of
$\chibh$ significantly only if signal $\rho\gtrsim 30$. 
%
Overall, we learn an important thing. Bayesian parameter estimation algorithms
aimed at identifying NSBH binaries with low latency in order to alert 
electromagnetic (EM) observers in time to establish a possible coincident 
observation of a GRB~\cite{2012A&A...541A.155A,Singer:2014qca,Singer:2015ema,
Pankow:2015cra} can use BBH templates to look for NSBH systems with aLIGO.
This is so because the primary requirement of the EM follow-up effort is rapid
classification of a binary as an NSBH system, which can be achieved just as 
easily with BBH templates, purely on the basis of the smaller component's mass.
Another conclusion we draw is that detection searches are unaffected by the
choice of ignoring tidal effects in matched-filtering templates.

@@

Multiple realistic observations deliver far more information than what a single
observation furnishes, as has been shown the case to be for binary neutron
stars~\cite{DelPozzo:13}. As our final study, we perform a fully-Bayesian 
analysis of how our estimation of $\lambdans$ changes as we accumulate more and
more detections. The choices that determine the population are important here,
and were made as follows. Neutron star mass is fixed for signals at
$1.35M_\odot$, and its spin $\chins=0$. Black hole mass is sampled uniformly
from the range $[2,5]\times 1.35=[2.7, 6.75]M_\odot$. Our choice here is given
by the intersection set of the mass range that allows for neutron star disruption
and the range supported by the waveform model LEA+~\cite{Foucart2012,
Foucart:2013a,Lackey:2013axa}. Black hole spin is sampled uniformly from the
{\it aligned} range, i.e. $\chibh\in[0,+1]$. The true deformability is fixed
once per population, sampled uniformly from the range $\lambdans\in[0,2000]$.
In order to keep the computational cost tractable, we make a further 
approximation. We inject signals into zero noise at different SNRs and recover
their parameters from Bayesian inferencing for a fixed set of binaries,
whose parameters lie on a uniform grid given by:
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
Once the population parameters been selected, each event's parameters are
replaced by their nearest neighbor in the above set. We make sure that this 
approximation is conservative by keeping our population sampling ranges within the
conservative side of the range spanned by the above set. With the caveats
laid out, we report our results in Sec.~\ref{s1:multiple_observations}.
In addition, we probe two paradigms, one that allows for BH masses to lie
within the astrophysical mass-gap (paradigm A), and one that does not (paradigm
B).
% 
For paradigm A, we find the following: (i) for the softer equations of state
that result in less deformable neutron stars, $15-20$ detections are sufficient
for our median measured value of $\lambdans$ to lie within $10\%$ of the true
value. (ii) For NSBH populations with more deformable NSs
($\lambdans\geq 1000$), the same is achievable within as few as $10$ realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans\geq 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs do indeed have
masses between $\leq 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
% 
For paradigm B, we find the information accumulation to be somewhat slower.
While the quantitative inferences for $\lambdans\geq1000$ populations are
not affected significantly, we find $\mathcal{O}(10-20)\%$ more events are
required to attain the same measurement accuracy as under paradigm A.
% 
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, with a few years'
time-scale.

>>>>

\author{Prayush Kumar}\CITA\email{prkumar@cita.utoronto.ca}
\author{Michael P\"urrer}\AEI
\author{Harald P. Pfeiffer}\CITA\CIFAR\AEI

@@

\begin{abstract}
The pioneering discovery of gravitational waves (GW) by Advanced LIGO has ushered
us into an era of observational GW astrophysics. Compact binaries remain the 
primary target sources for GW observation, of which neutron star - black hole 
(NSBH) binaries form an important subset.
% 
GWs from NSBH sources carry signatures of (a) the star's tidal distortion by
the companion black hole (BH) during inspiral, and (b) its tidal disruption
(if that happens) near merger.
%
In this paper, we present the first fully Bayesian study of the measurability 
of neutron star (NS) tidal deformability $\lambdans\propto (R/M)_\mathrm{NS}^{5}$ from NSBH
observation(s), where we
take into account the crucial effect of BH spins as well.
% 
First, we find that if BHBH templates are used to estimate source 
parameters for an NSBH signal, a significant bias is introduced in the 
inference if the signal-to-noise ratio (SNR) is greater than $\simeq30$.
% 
Second, we find that if we are to get a loud signal with SNR $\gtrsim 23$, 
we can put a factor of two bound on $\lambdans$.
% 
Finally, we study the improvement in our measurement of $\lambdans$ from 
multiple observations. Using an approximately astrophysical population, 
we find that (i) the median measured $\lambdans$ catches up to within
$10\%$ of the true value in approximately $20$ observations; (ii) in the same
number of observations, the associated $90\%$ confidence intervals shrink to
$\pm 50\%$ of the true value.
% 
We find these results encouraging, and recommend that an effort to 
measure $\lambdans$ be taken within the LIGO-Virgo Collaboration once
we begin to detect NSBH coalescences.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{s1:introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 
% Over the $3$-months of O1, aLIGO had access to more binary mergers than 
% initial-LIGO over its entire coincident operation period.

The Advanced LIGO (aLIGO) observatories completed their first observing run ``O1''
early-2016, operating at a factor of $3-4$ higher gravitational-wave
(GW) strain sensitivity than their first-generation 
counterparts~\cite{Shoemaker2009}.
% 
During O1, aLIGO made the first terrestrial observation of gravitational 
waves~\cite{LIGOVirgo2016a}. Emitted by a pair of coalescing black holes, these
waves
heralded an era of observational GW astrophysics as they traveled through Earth.
% 
By $2018-19$, we expect the aLIGO instruments to reach their design
sensitivity, at which point we may expect to observe $\sim 70$ binary mergers a
year~\cite{Abadie:2010cf}.

Coalescing binaries of compact objects, i.e. black holes (BH) and/or
neutron stars (NS) are the primary target sources for 
aLIGO~\cite{Abbott:2016nhf}. An important sub-class of these sources are NSBH
binaries, of which we expect $\mathcal{O}(10)$ GW observations every 
year~\cite{Abadie:2010cf}.
NSBH binaries are of interest for multiple reasons. For instance,
they have been long associated with (as possible progenitors of) short
Gamma-ray Bursts (GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Under favorable conditions, the NS can get disrupted by
the tidal field of its companion BH. Once disrupted, most of the NS
material falls into the BH over an $\mathcal{O}(1$ms$)$ time-scale,
with the rest forming an accretion disk around the BH. This disk-BH system,
which survives for $0.1-1$s, is hypothesized to drive SGRBs through the
production of relativistic jets~\cite{Foucart2015,
Lovelace:2013vma,Deaton2013,Foucart2012,Shibata:2005mz,Paschalidis2014}.
A {\it coincident} detection of both gravitational and electromagnetic
waves (gamma rays) from an NSBH system is therefore a {\it unique}
possibility to confirm this hypothesis~\cite{Abbott:2016wya}.
% 
% Most of the matter falls into the BH in ~1ms. But the disk itself has an
% accretion timescale of ~100ms-1s. The disk-BH system is the expected engine
% for a SGRB, through the production of a relativistic jet. The unbound material,
% on the other hand, is responsible for an EM signal in the infrared/optical, 
% powered by radioactive decays in the ejecta. And possibly a radio afterglow
% through interactions with the ISM. But not gamma-rays (unless you are referring
% to the very small amount of material ejected in the jet, but that's not what
% people usually mean when talking about material unbound by the merger).
% 
Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron star material. This is what we will focus on in this paper.
% 
GWs emitted by NSBH coalescences carry subtle hints of the material structure
of NSs. During early inspiral, the (weak) tidal field of the companion BH
deforms the star, thereby exciting its resonant oscillation modes. The induced
quadrupolar moment $Q_{ij}$ of the star depends on the
tidal field $\mathcal{E}_{ij}$ of its companion through the relation
$Q_{ij} = -\lambda\mathcal{E}_{ij}$, where $\lambda$ is an EoS-dependent
tidal deformability parameter that is related to NS's dimensionless Love number
$k_2$ and radius $R_\mathrm{NS}$ as $\lambda:=\frac{2}{3G}k_2 R_\mathrm{NS}^5$.
The coherence of GW emission between stellar quadrupolar waves and orbital
waves drains energy from the orbit and increases the inspiral rate~\cite{Flanagan2008}.
We know the leading and next-to
-leading order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN order. As the NS gets
disrupted close to binary's merger, the quadrupole moment of the system falls
monotonically. As a result, the emitted GWs are subdued significantly
post-disruption (as compared to what they would be for a 
BBH)~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,Foucart2015,
Pannarale:2015jia}.

The difficulty in discerning NS matter effects in an NSBH coalescence from its
GW signal lies in distinguishing it reliably from a BBH signal, since the 
difference between the two is either subtle (during inspiral) or short-lived
(disruption and merger). The earlier in inspiral the NS disrupts, the more
orbits there are over which its effect is visible. The mechanism of disruption
is governed by the nature of the BH, i.e. its mass and spin. More massive BHs 
(with $m_\mathrm{BH}\gtrsim 7\, m_\mathrm{NS}$), as well as BHs with
(large) retrograde spins $(\chi_\mathrm{BH}\in[-1. 0])$, tend to swallow the 
NS whole due to a rapid merger (i.e. without it disrupting before the
inner-most stable circular orbit (ISCO) is reached). The quasi-normal
oscillation modes of the final black hole's horizon are also 
excited~\cite{Foucart:2013psa}. Therefore, the only available information 
comes from the inspiral rate modification. On the other hand, less massive
black holes spinning in the prograde sense (with mass-ratio 
$q\in[2, 5]$, and $\chi_\mathrm{BH}\geq 0$) can cause their companion star to 
disrupt much before merger, leaving a strong imprint on the emitted GW 
signal~\cite{Shibata:2007zm,2010PhRvD..81f4026F,Lovelace:2013vma,
Kawaguchi:2015}.

@@

We first probe the effect of not including tidal effects in parameter 
estimation templates on the recovery of non-tidal binary parameters, such as 
component masses and spins. This is the case of current and planned aLIGO
efforts.
To do so, we use the LEA+ model to generate a set of realistic signals, and
use that as data from which we estimate binary mass and spin parameters with
a Bayesian Monte-Carlo algorithm, using non-tidal (BBH) waveforms as filters.
% 
As in rest of the paper, we restrict here our parameter space to span 
mass-ratios $q:=\mbh/\mns\in[2,5]$, black hole spin (aligned with orbit)
$\chibh\in[-0.5, +0.75]$, and dimensionless neutron star tidal deformability 
$\lambdans:= G\left(\frac{c^2}{G \mns}\right)^5\lambda \in[500, 2000]$.
The mass of NS for signals (and not templates) is fixed to $1.35M_\odot$ 
throughout~\cite{stellarcollapsemass}, and is taken to be 
non-spinning~\cite{Miller:2014aaa}.
Each signal is rescaled to a range of SNR values $\rho\in[10, 70]$.
% 
We find that non-inclusion of tidal effects leads to systematic biases in
the measurement of non-tidal parameters. Consider first the binary chirp 
mass $\mchirp:=M\eta^{3/5}$ where $M$ is the binary total mass, and 
$\eta:=\mbh\mns/M^2$ is the dimensionless mass-ratio. We find that for
realistic (moderate) SNRs, i.e. $\rho\lesssim 30$, not including tidal
effects introduces a barely measurable bias that remains below $50\%$
of the statistical uncertainty of the $\mchirp$ measurement itself.
For higher signal strengths, we find that the underlying statistical
uncertainty decreases enough for the systematic bias to become comparable
to it for $\rho=50-70$ and higher. This is expected because chirp mass is fairly
accurately determined by the inspiral, where tidal effects are weak.
Next, we consider the mass-ratio $\eta$. We find that only if the NS is quite
deformable (with $\lambdans=2000$), {\it and} has a companion BH whose mass
is $\mbh\lesssim 4.5M_\odot$ (i.e. in the astrophysical ``mass-gap''~\cite{
Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa}),
will the systematic bias in $\eta$ measurement become comparable to its
statistical uncertainties at SNRs $\rho\leq 30$. For SNRs $\rho\gtrsim30$,
however, not including tidal effects is very likely to {\it significantly} bias
the mass-ratio estimate for NSBH binaries. Finally, we consider the spin on the
black hole $\chibh$. We find that $\chibh$ measurements are affected in a similar
way to $\eta$. By not including tidal effects, we compromise our measurement of
$\chibh$ significantly only if signal $\rho\gtrsim 30$. 
%
Overall, we learn an important thing. Bayesian parameter estimation algorithms
aimed at identifying NSBH binaries with low latency in order to alert 
electromagnetic (EM) observers in time to establish a possible coincident 
observation of a GRB~\cite{2012A&A...541A.155A,Singer:2014qca,Singer:2015ema,
Pankow:2015cra} can use BBH templates to look for NSBH systems with aLIGO.
This is so because the primary requirement of the EM follow-up effort is rapid
classification of a binary as an NSBH system, which can be achieved just as 
easily with BBH templates, purely on the basis of the smaller component's mass.
Another conclusion we draw is that detection searches are unaffected by the
choice of ignoring tidal effects in matched-filtering templates.

@@

Multiple realistic observations deliver far more information than what a single
observation furnishes, as has been shown the case to be for binary neutron
stars~\cite{DelPozzo:13}. As our final study, we perform a fully-Bayesian 
analysis of how our estimation of $\lambdans$ changes as we accumulate more and
more detections. The choices that determine the population are important here,
and were made as follows. Neutron star mass is fixed for signals at
$1.35M_\odot$, and its spin $\chins=0$. Black hole mass is sampled uniformly
from the range $[2,5]\times 1.35=[2.7, 6.75]M_\odot$. Our choice here is given
by the intersection set of the mass range that allows for neutron star disruption
and the range supported by the waveform model LEA+~\cite{Foucart2012,
Foucart:2013a,Lackey:2013axa}. Black hole spin is sampled uniformly from the
{\it aligned} range, i.e. $\chibh\in[0,+1]$. The true deformability is fixed
once per population, sampled uniformly from the range $\lambdans\in[0,2000]$.
In order to keep the computational cost tractable, we make a further 
approximation.
%
\prayush{%
We inject signals into zero noise at different SNRs and recover
their parameters from Bayesian inferencing for a fixed set of binaries,
whose parameters lie on a uniform grid given by:
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
% 
Once the population parameters been selected, each event's parameters are
replaced by their nearest neighbor in the above set.
}
We make sure that this approximation is conservative by keeping our population
sampling ranges within the conservative side of the range spanned by the above
set. With the caveats laid out, we report our results in
Sec.~\ref{s1:multiple_observations}. In addition, we probe two paradigms, one
that allows for BH masses to lie within the astrophysical mass-gap (paradigm
A), and one that does not (paradigm B).
% 
For paradigm A, we find the following: (i) for the softer equations of state
that result in less deformable neutron stars, $15-20$ detections are sufficient
for our median measured value of $\lambdans$ to lie within $10\%$ of the true
value. (ii) For NSBH populations with more deformable NSs
($\lambdans\geq 1000$), the same is achievable within as few as $10$ realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans\geq 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs do indeed have
masses between $\leq 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
% 
For paradigm B, we find the information accumulation to be somewhat slower.
While the quantitative inferences for $\lambdans\geq1000$ populations are
not affected significantly, we find $\mathcal{O}(10-20)\%$ more events are
required to attain the same measurement accuracy as under paradigm A.
% 
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, with a few years'
time-scale.
